---
title: "Reanalysis of Libro, S., Characterizing innate immunity, PRJNA329497"
author: "Christopher Holt"
date: "23-May-2024"
output: pdf_document
---

# Load Libraries

```{r Load in libraries}
library(edgeR)
library(FactoMineR)
library(ggdendro)
library(tidyverse)
library(pvclust)
library(variancePartition)
library(WGCNA)
library(devtools)
devtools::source_url("https://raw.githubusercontent.com/obigriffith/biostar-tutorials/master/Heatmaps/heatmap.3.R")
```

# Session Info

```{r print session info}
sessionInfo()
```

# Load in Functions

```{r load functions}
## Add mitochondrial gene information to GENEINFO
add_mitochondrial_genes <- function(df, column_name){
  df <- df %>% mutate("{column_name}" := ifelse(grepl('gene11440', df[[column_name]]), 'gene-nad2',
                                                ifelse(grepl('gene11441', df[[column_name]]), 'gene-nad4',
                                                       ifelse(grepl('gene11442', df[[column_name]]), 'gene-cox1',
                                                              ifelse(grepl('gene11443', df[[column_name]]), 'gene-nad6',
                                                                     ifelse(grepl('gene11444', df[[column_name]]), 'gene-cob',
                                                                            ifelse(grepl('gene11445', df[[column_name]]), 'gene-cox3',
                                                                                   ifelse(grepl('gene11446', df[[column_name]]), 'gene-nad4L',
                                                                                          ifelse(grepl('gene11447', df[[column_name]]), 'gene-nad1',
                                                                                                 ifelse(grepl('gene11448', df[[column_name]]), 'gene-atp6',
                                                                                                        ifelse(grepl('gene11449', df[[column_name]]), 'gene-cox2',
                                                                                                               ifelse(grepl('gene11450', df[[column_name]]), 'gene-nad3',
                                                                                                                      ifelse(grepl('gene11451', df[[column_name]]), 'gene-nad5', df[[column_name]])))))))))))))


  return(df)

}

create_matrix <- function(df, row, centre, column){
  df <- as.data.frame(df)
  test <- df %>% select(all_of(row), all_of(centre), all_of(column))
  replicates.test <- unique(test[[column]])
  tmp.matrix <- tibble(gene=unique(df[[row]])) ## row
  for (replicate_uniq in replicates.test){
    tmp <- test %>% filter(!!as.symbol(column) == replicate_uniq)
    colnames(tmp)[2] <- replicate_uniq
    tmp[[column]] <- NULL
    tmp.matrix <- inner_join(tmp.matrix, tmp, by = c('gene' = row))
  }

  tmp.matrix <- tmp.matrix %>% tibble::column_to_rownames('gene')
  return(tmp.matrix)

}


calculate_TPM <- function(counts.mat, counts.orig){
  gene_length <- counts.orig %>% select('Name', 'length') %>% unique
  for (i in 1:ncol(counts.mat)){
    counts.mat[,i] <- counts.mat[,i]/gene_length[,2]
    counts.mat[,i] <- counts.mat[,i]/(sum(counts.mat[,i])/1000000)
  }

  return(counts.mat)
}

find_soft_power <- function(sft){
  df <- as.data.frame(cbind(sft$fitIndices[,1], -sign(sft$fitIndices[,3])*sft$fitIndices[,2]))
  y <- -sign(sft$fitIndices[,3])*sft$fitIndices[,2]
  dy <- diff(y) 
  softpower <- which(abs(diff(y)) < 0.05)[1]
  return(softpower)
}

eigengene_invert_id <- function(tpm.de.wgcna, mergedColors, mergedMEs){
  tpm.de <- tpm.de.wgcna
  tpm.de.wgcna$invert <- T
  tpm.de.wgcna$module <- mergedColors
  for(i in 1:nrow(tpm.de.wgcna)){
    if(cor(t(tpm.de[i,]), mergedMEs[,which(colnames(mergedMEs) == paste0("ME",tpm.de.wgcna$module[i]))], method = "pearson") >= 0){
      tpm.de.wgcna$invert[i] <- F
    }
  }
  return(tpm.de.wgcna)
}

wgcna_heatmap_reorder <- function(tpm.de.wgcna){
  clusters <- as.data.frame(table(tpm.de.wgcna$module))
  clusters <- clusters[order(-clusters[,2]),1]

  tpm.de.wgcna.reordered <- as.data.frame(matrix(nrow = 0,
                                                 ncol = ncol(tpm.de.wgcna)))
  for(i in 1:length(clusters)){
    tpm.de.wgcna.reordered <- as.data.frame(rbind(tpm.de.wgcna.reordered,
                                                  tpm.de.wgcna[tpm.de.wgcna$module == clusters[i] & tpm.de.wgcna$invert == F,],
                                                  tpm.de.wgcna[tpm.de.wgcna$module == clusters[i] & tpm.de.wgcna$invert == T,]))
  }
  return(tpm.de.wgcna.reordered)
}

get_heatmap_separators <- function(vector){
  sep <- c()
  for(i in 2:length(unique(vector))){
    sep[length(sep) + 1] <- min(which(vector == unique(vector)[i])) - 1
  }
  return(sep)
}


eigengene_refit <- function(tpm.de, mergedMEs, min_pearson_corr = 0.8){
  tpm.de.wgcna <- tpm.de
  tpm.de.wgcna$invert <- T
  tpm.de.wgcna$module <- NA
  for(i in 1:nrow(tpm.de)){
    minimum_correlation <- min_pearson_corr
    for(j in 1:ncol(mergedMEs)){
      cor <- abs(cor(t(tpm.de[i,]), mergedMEs[,j], method = "pearson"))
      if(cor > minimum_correlation){
        minimum_correlation <- cor
        tpm.de.wgcna$module[i] <- colnames(mergedMEs)[j]
      }
    }
    if(!is.na(tpm.de.wgcna$module[i])){
      if(cor(t(tpm.de[i,]), mergedMEs[,which(colnames(mergedMEs) == tpm.de.wgcna$module[i])], method = "pearson") > 0){
        tpm.de.wgcna$invert[i] <- F
      }
    }else{
      tpm.de.wgcna$invert[i] <- NA
    }

  }
  tpm.de.wgcna$module <- gsub("^ME","",tpm.de.wgcna$module)
  return(tpm.de.wgcna)
}

consolidate_functional_terms <- function(geneinfo){

  ipr <- as.data.frame(unlist(strsplit(paste(geneinfo$interpro_description, collapse = "|"),  split = "[|]")))
  gobio <- as.data.frame(unlist(strsplit(paste(geneinfo$go_biologicalprocess, collapse = "|"),  split = "[|]")))
  gocell <- as.data.frame(unlist(strsplit(paste(geneinfo$go_cellularcomponent, collapse = "|"),  split = "[|]")))
  gomol <- as.data.frame(unlist(strsplit(paste(geneinfo$go_molecularfunction, collapse = "|"),  split = "[|]")))

  ipr <- as.data.frame(table(ipr))
  gobio <- as.data.frame(table(gobio))
  gocell <- as.data.frame(table(gocell))
  gomol <- as.data.frame(table(gomol))

  functionalterms.list <- list(ipr = ipr, gobio = gobio, gocell = gocell, gomol = gomol)

  return(functionalterms.list)
}


functionaltermenrichment <- function(genes, geneinfo, functionalterms.list){

  geneinfo.subset <- geneinfo[geneinfo$gene %in% genes,]
  term <- c()
  clusteroccurences <- c()
  genomeoccurences <- c()
  pvalue <- c()
  correctedpvalue <- c()
  oddsratio <- c()
  clusterProportion <- c()
  genomeProportion <- c()
  clusterSize <- c()
  genomeSize <- c()
  newOddsRatio <- c()

  functionalterms.list.subset <- consolidate_functional_terms(geneinfo.subset)

  for(i in 1:length(functionalterms.list)){
      # pvalue_cluster <- c()
      # corrected_pvalue_cluster <- c()
    for(j in 1:nrow(functionalterms.list[[i]])){

      freq.all <- functionalterms.list[[i]][j,2] ## How amany times the IPR/GO term appears in the entire geneinfo file

      ## determine if the functional term is in the subset
      ## If it isn't, then the frequency is set to 0
      ## If it is in the subset, then extract the frequency of the IPR/GO term from within the subset
      freq.subset <- ifelse(functionalterms.list[[i]][j,1] %in% functionalterms.list.subset[[i]][,1],
                            functionalterms.list.subset[[i]][functionalterms.list.subset[[i]][,1] == as.character(functionalterms.list[[i]][j,1]),2],
                            0)

      ## Number of genes in the whole genome
      genes.all <- nrow(geneinfo)

      ## Number of genes in the cluster
      genes.subset <- nrow(geneinfo.subset)
      # [col.row]
      ## [1,1] is frequency is the subset, [1,2] is remaining frequency in rest of gene term file (col 1 total is total frequency of term in genome)
      ## [2,1] is no. genes that do not have this term in the subset, [2,2] is no. genes that do not have this term in genome (minus genes in subset)
      fisherexact.matrix <- matrix(c(freq.subset, freq.all - freq.subset,
                                     genes.subset - freq.subset, genes.all - genes.subset - freq.all + freq.subset),
                                   nrow = 2,
                                   ncol = 2)
      fisher.test <- fisher.test(fisherexact.matrix)

      term[length(term) + 1] <- as.character(functionalterms.list[[i]][j,1])
      clusteroccurences[length(clusteroccurences) + 1] <- as.numeric(as.character(freq.subset))
      genomeoccurences[length(genomeoccurences) + 1] <- as.numeric(as.character(freq.all))
      pvalue[length(pvalue) + 1] <- as.numeric(as.character(fisher.test$p.value))
      correctedpvalue[length(correctedpvalue) + 1] <- p.adjust(as.numeric(as.character(fisher.test$p.value)), method = "fdr", n = nrow(functionalterms.list[[i]]))
      oddsratio[length(oddsratio) + 1] <- as.numeric(as.character(fisher.test$estimate)) ## if < 1, term is under represented in the clutser

      clusterProportion[length(clusterProportion) + 1] <- as.numeric(fisherexact.matrix[1,1]/fisherexact.matrix[2,1])
      clusterSize[length(clusterSize) + 1] <- as.numeric(fisherexact.matrix[2,1])
      genomeProportion[length(genomeProportion) + 1] <- as.numeric(fisherexact.matrix[1,2]/fisherexact.matrix[2,2])
      genomeSize[length(genomeSize) + 1] <- as.numeric(fisherexact.matrix[2,2])

      newOddsRatio[length(newOddsRatio) + 1] <- as.numeric((fisherexact.matrix[1,1]/fisherexact.matrix[2,1])/(fisherexact.matrix[1,2]/fisherexact.matrix[2,2]))

    }
  }

  terms.df <- as.data.frame(cbind(term,
                                  clusteroccurences,
                                  genomeoccurences,
                                  pvalue,
                                  correctedpvalue,
                                  oddsratio,
                                  clusterProportion,
                                  clusterSize,
                                  genomeProportion,
                                  genomeSize,
                                  newOddsRatio))
  terms.df <- terms.df[order(as.numeric(as.character(terms.df$pvalue))),]
  return(terms.df)
}
  
ggplot_legend <- function(a.gplot){ 
  tmp <- ggplot_gtable(ggplot_build(a.gplot)) 
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box") 
  legend <- tmp$grobs[[leg]] 
  return(legend)

}
tidy_matrix <- function(df){
  tidy_df <- tibble()
  cols <- colnames(df %>% select(-'invert', -'module'))
  for(col in cols){
    tmp <- df %>% select(col)
    colnames(tmp)[1] <- 'z_score'
    tmp$replicate <- col
    tmp <- tmp %>% rownames_to_column('gene')
    tmp <- tmp %>% select('gene', 'replicate', 'z_score')
    tidy_df <- rbind(tidy_df, tmp)
  }
  
  tidy_df <- inner_join(tidy_df,
                        df %>%
                          rownames_to_column('gene') %>%
                          select('gene', 'invert', 'module'),
                        by = 'gene')
  
  return(tidy_df)
}


factor_groups <- function(GROUPS){
  
  for (col in colnames(GROUPS)){
    GROUPS[[col]] = factor(GROUPS[[col]], levels = unique(GROUPS[[col]]))
  }
  
  return(GROUPS)
}


identify_de_genes <- function(counts.mat, groups.df, returnDGE = FALSE){
  FDRcutoff <- 0.05
  #FDRcutoff <- 0.2
  cpm.cutoff <- 5/min(colSums(counts.mat)) * 1000000 ## list minimum counts per million
  print(paste0(nrow(counts)," total genes"))
  y <- edgeR::DGEList(counts = counts.mat, group = groups.df$grouping_factor)
  y <- edgeR::calcNormFactors(y)
  keep <- rowSums(cpm(y) >= cpm.cutoff) >= min(table(groups.df$grouping_factor)) ## is the cpm > than cutoff in two or more replicates
  keep.df <- as.data.frame(table(keep))
  print(paste0(keep.df[keep.df[,1] == F,2]," genes excluded using edgeR filter"))
  y <- y[keep, , keep.lib.sizes = F] ## Selects genes in y that are in keep
  design <- model.matrix(~groups.df$grouping_factor)
  y <- estimateDisp(y , design)
  ## Next two steps use Genewise negative binomial generalised linear models with quasilikelihood tests
  fit <- edgeR::glmQLFit(y, design) 
  qlf <- edgeR::glmQLFTest(fit, coef = 2:ncol(fit)) 
  FDR <- as.data.frame(p.adjust(qlf$table$PValue, method="BH")) ## Takes pvalue from above and correct using FDR method
  rownames(FDR) <- rownames(qlf$table)
  colnames(FDR) <- FDR
  degenes <- rownames(FDR[FDR[,1] < FDRcutoff, , drop = F]) ## Select genes with FDR less than FDR cutoff
  print(paste0(length(degenes)," DE genes"))
  
  if(returnDGE){
    return(y)
  }else{
    return(degenes)
  }
}

plot_de_genes <- function(tpm, degenes, groups.df, OUTFILE, column_colour, y_dge = NULL, wgcna_clustering = TRUE, z_score = FALSE){
  
  if(wgcna_clustering){
    tpm.de <- tpm[rownames(tpm) %in% degenes,colnames(tpm) %in% groups.df$Replicate]
    tpm.de <- tpm.de[,levels(groups.df$Replicate)]
    
    wgcna <- as.data.frame(t(tpm.de))
    powers <- c(c(1:10), seq(from = 12, to=20, by=2))
    sft <- pickSoftThreshold(wgcna, powerVector = powers, verbose = 5)
    softpower <- find_soft_power(sft)
    
    adjacency <- adjacency(wgcna, power = softpower)
    TOM <- TOMsimilarity(adjacency)
    dissTOM <- 1-TOM
    geneTree <- hclust(as.dist(dissTOM), method = "average");
    minModuleSize <- 1
    dynamicMods <- cutreeDynamic(dendro = geneTree, distM = dissTOM,
                                 deepSplit = 2, pamRespectsDendro = FALSE,
                                 minClusterSize = minModuleSize)
    dynamicColors = labels2colors(dynamicMods)
    MEList = moduleEigengenes(wgcna, colors = dynamicColors)
    MEs = MEList$eigengenes
    MEDiss = 1-cor(MEs, use = "pairwise.complete.obs")
    METree = hclust(as.dist(MEDiss), method = "average")
    MEDissThres = 0.25
    
    
    merge = mergeCloseModules(wgcna, dynamicColors, cutHeight = MEDissThres, verbose = 3)
    mergedColors = merge$colors
    mergedMEs = merge$newMEs
    
    tpm.de.wgcna <- eigengene_invert_id(tpm.de, mergedColors, mergedMEs)
    tpm.de.wgcna <- wgcna_heatmap_reorder(tpm.de.wgcna)
    
    log2tpm.de <- log2(tpm.de.wgcna[,1:(ncol(tpm.de.wgcna) - 2)] + 1)
    zscore.log2tpm.de <- as.data.frame(t(scale(t(log2tpm.de))))
    
    ##-----------------------------------------------------------------------------------------------------------
    ## Add invert and module column to the zscore dataframe
    ##-----------------------------------------------------------------------------------------------------------
    
    zscore.log2tpm.de <- inner_join(zscore.log2tpm.de %>% rownames_to_column('gene'), 
               tpm.de.wgcna %>% select('invert', 'module') %>% rownames_to_column('gene'), by = 'gene') %>% column_to_rownames('gene')
    
    
    # zscore.log2tpm.de <- zscore.log2tpm.de %>% filter(rownames(.) == "WBGene00222154" | rownames(.) == "WBGene00227885")  
    # tpm.de.wgcna <- tpm.de.wgcna %>% filter(rownames(.) == "WBGene00222154" | rownames(.) == "WBGene00227885")  
    
    
    hmcol <- colorRampPalette(c("navyblue","white","firebrick3"))(12)
    rowcol <- tpm.de.wgcna$module
    colcol <- as.character(groups.df$column_colour)
    rowsep <- get_heatmap_separators(rowcol)
    colsep <- get_heatmap_separators(colcol)
    breaks <- seq(-3,3,by=.5)
    
    rowcol1 <- zscore.log2tpm.de$invert
    rowcol1[rowcol1 == F] <- "grey" ## matching pattern
    rowcol1[rowcol1 == T] <- "black" ## inverted pattern
    
    rowcol2 <- t(cbind(rowcol, rowcol1))
    
    
    ##-----------------------------------------------------------------------------------------------------------
    ## Set the row colours and coloumn colours as matrices
    ##-----------------------------------------------------------------------------------------------------------
    
    rowcol2 <- as.matrix((rowcol2))
    colcol <- as.matrix(colcol)
    
    heatmap_matrix <- as.matrix(zscore.log2tpm.de[,(1:(ncol(zscore.log2tpm.de)-2))])
    
    Rowv_clustering = F
    Colv_clustering = F
    RowColourSize = 1.5

    
    RowColours = rowcol2
    
    dend = "none"
    
  }else{
    
    if(!is.null(y_dge)){
        tpm.keep <- tpm[rownames(tpm) %in% rownames(y_dge$counts),colnames(tpm) %in% groups.df$Replicate]
        tpm.keep <- tpm.keep[,levels(groups.df$Replicate)]

    }else{
        tpm.keep <- tpm[rownames(tpm) %in% degenes,colnames(tpm) %in% groups.df$Replicate]
        tpm.keep <- tpm.keep[,levels(groups.df$Replicate)]

    }
    
    if(z_score){
      log2.tpm.keep <- log2(tpm.keep+1)
      zscore.log2.tpm.keep <- as.data.frame(t(scale(t(log2.tpm.keep))))
      zscore.log2.tpm.keep <- zscore.log2.tpm.keep[,levels(groups.df$Replicate)]


      breaks <- seq(-3,3,by=.5)
      hmcol <- colorRampPalette(c("navyblue","white","firebrick3"))(12)
      
      zscore.log2.tpm.keep$module <- 'white'
      RowColours = as.matrix(t(zscore.log2.tpm.keep$module))


      heatmap_matrix <- zscore.log2.tpm.keep[,(1:(ncol(zscore.log2.tpm.keep)-1))]

    }else{
      log2.tpm.keep <- log2(tpm.keep+1)
      breaks <- seq(min(log2.tpm.keep),max(log2.tpm.keep),by=0.5)
      hmcol <- colorRampPalette(c("white","firebrick1","firebrick3"))(length(breaks)-1)
    
      log2.tpm.keep$module <- 'white'
      RowColours = as.matrix(t(log2.tpm.keep$module))


      heatmap_matrix <- log2.tpm.keep[,(1:(ncol(log2.tpm.keep)-1))]
      heatmap_matrix <- heatmap_matrix[,levels(groups.df$Replicate)]

    }

    
    rowsep = FALSE
    colcol <- as.matrix(as.character(groups.df[[column_colour]]))
    colsep = FALSE
    Rowv_clustering = T
    Colv_clustering = F
    RowColourSize = 0.5
    dend = "row"
  }
  
  
  heatmap_matrix <- heatmap_matrix[,levels(groups.df$Replicate)]
  
  
  pdf(paste0(FIGURE_DIR, "/", OUTFILE, ".pdf"),
      width = 10,
      height = 10)
  heatmap.3(heatmap_matrix,
                col=hmcol,
                trace="none",
                labRow = NA,
                labCol = F,
                Rowv = Rowv_clustering,
                Colv = Colv_clustering,
                RowSideColors=RowColours,
                RowSideColorsSize = RowColourSize,
                ColSideColors=colcol,
                lhei = c(2,8,1),
                breaks = breaks,
                rowsep = rowsep,
                density.info = 'histogram',
                dendrogram = dend)
  
  invisible(dev.off())
  
}

variance_function <- function(counts.to.use, groups.to.use, form){
  y <- edgeR::DGEList(counts = counts.to.use)
  y <- edgeR::calcNormFactors(y)
  design <- stats::model.matrix(~groups.to.use$Sample)
  
  y <- limma::voom(y, design)

  varPart <- fitExtractVarPartModel(y, form, groups.to.use)
  
  varPart.df <- gg_plotvar(varPart)
  
  return(varPart.df)
}

gg_plotvar <- function(varPart){
  varPart.df <- data.frame()
  varPart <- as.data.frame(varPart) %>% rownames_to_column('gene')
  for(col in colnames(varPart)[colnames(varPart) != 'gene']){
    df <- as.data.frame(varPart) %>% select('gene',all_of(col))
    df$attribute <- col
    colnames(df)[2] <- 'Variance_Explained'
    
    varPart.df <- rbind(varPart.df, df)
  }
  
  
  varPart.df$attribute <- factor(varPart.df$attribute, levels = unique(varPart.df$attribute))

  
  return(varPart.df)

}

median_variance <- function(varPart.df, attr_list){
  
  for(attr in attr_list){
    message(paste0(attr))
      print(
    median(varPart.df[varPart.df$attribute == paste0(attr),]$Variance_Explained)
  )
  }
}
remove_all_zero_rows <- function(df){
  keep = rowSums(df) > 0
  df.keep <- df[keep,]
  return(df.keep)
}




plot_pca <- function(tpm.to.use, groups.df, OUTFILE, counts){
  pca <- PCA(t(as.data.frame(t(scale(t(log2(tpm.to.use+1)))))), graph = FALSE, ncp = ncol(counts) - 1)
  pdf(paste0(FIGURE_DIR, "/", OUTFILE, ".pdf"),
       width = 10,
       height = 10)
    print(
      ggplot()+
        geom_point(aes(x=pca$ind$coord[,1], y=pca$ind$coord[,2], color = groups.df$Sample, size = 3))+
        labs(col = "Samples", size = "Reads Mapped\nto Features", 
             x = paste0("PC1 (", round(pca$eig[1,2],1), "%)"), 
             y = paste0("PC2 (", round(pca$eig[2,2],1), "%)"))+
        guides(colour = guide_legend(ncol = 2), size = 'none')+
        scale_color_manual(values = levels(groups.df$column_colour))+
        theme_bw()
    )
  dev.off()

}

plot_dendrogram <- function(tpm.to.use, groups.df, OUTFILE){
  
  tpm_dendrogram <- pvclust::pvclust(as.data.frame(t(scale(t(log2(tpm.to.use+1))))), method.dist="cor", method.hclust="average",
                                   nboot=1000, parallel=TRUE)

  pdf(paste0(FIGURE_DIR, "/", OUTFILE, ".pdf"),
        width = 10,
        height = 10)
    plot(tpm_dendrogram)
    
  invisible(dev.off())
  
  dendro.data <- ggdendro::dendro_data(tpm_dendrogram$hclust)
  dendro_order <- dendro.data$labels %>% select('label')
  
  sample_order <- factor(as.character(dendro_order$label), levels = as.character(dendro_order$label))
  groups.df <- factor_groups(groups.df[order(match(groups.df$Replicate, as.character(sample_order))),])
  
  rownames(groups.df) <- NULL
  
  return(groups.df)

}


```

# Set Directory Paths

```{r set directory paths}
PROJECT="libro_PRJNA329497"
cwd <- getwd()
OUTPUT_DIR=paste0(cwd, "/output_files_", PROJECT)
FILE_DIR=paste0(OUTPUT_DIR, "/files_new")
FIGURE_DIR=paste0(OUTPUT_DIR, "/figures_new")

dir.create(OUTPUT_DIR, showWarnings = F)
dir.create(FILE_DIR, showWarnings = F)
dir.create(FIGURE_DIR, showWarnings = F)
```

# Set File Variables
```{r}
GENEINFO_URL <- 'https://raw.githubusercontent.com/Dunning-Hotopp-Lab/lf_transcriptome/master/input_data_files/bmalayi_gene.info'
ANNOTATION_URL <- "https://raw.githubusercontent.com/christopher-holt/bmalayi_meta_analysis/main/input_files/b_malayi.AF538716.AE017321.PRJNA10729.WS276.annotations.wormbase.gff3"
COUNTS_URL <- paste0('https://raw.githubusercontent.com/christopher-holt/bmalayi_meta_analysis/main/input_files/', PROJECT, '/combined.final.counts')
GROUPS_URL <- paste0('https://raw.githubusercontent.com/christopher-holt/bmalayi_meta_analysis/main/input_files/', PROJECT, '/groups.txt')
```



# Read in Files

```{r Read in Files}
# Read in geneinfo
geneinfo <- readr::read_delim(GENEINFO_URL, delim = '\t', show_col_types = F)

## Read in Groups File
groups <- utils::read.table(GROUPS_URL, sep = ',')

# Read in htseq counts and format into matrix
counts.orig <- utils::read.table(COUNTS_URL)


## Read in Annotation File
annotations <- readr::read_delim(ANNOTATION_URL, "\t", escape_double = FALSE, col_names = FALSE,
                                   trim_ws = TRUE, show_col_types = F)
```

# Format Groups File

```{r format groups}
groups <- as.data.frame(groups)
groups$V1 <- NULL
groups$Sample <- str_remove(string = groups$V2, pattern = '_[a-z]')
groups <- groups %>% mutate(column_colour = ifelse(groups$Sample == "Ba.24h", "red",
                                                   ifelse(groups$Sample == "Ba.36h", "darkred",
                                                          ifelse(groups$Sample == "CnoA.24h", "black",
                                                                 ifelse(groups$Sample == "CplusA.24h", "green",
                                                                        ifelse(groups$Sample == "CplusA.36h", "darkgreen",
                                                                               ifelse(groups$Sample == "DNA.12h", "blue",
                                                                                      ifelse(groups$Sample == "DNA.16h", "darkblue",
                                                                                             ifelse(groups$Sample == "Ec.20h", "yellow",
                                                                                                    ifelse(groups$Sample == "Ec.22h", "gold",
                                                                                                           ifelse(groups$Sample == "RNA.24h", "purple",
                                                                                                                  ifelse(groups$Sample == "RNA.36h", "violet", NA))))))))))))
colnames(groups) <- c("Replicate","Project", "Sample", "column_colour")
groups <- factor_groups(groups)
```



# Calculate Gene Lengths
```{r}

annotations <- annotations %>% separate('X9', into = c("Name", "X9"), sep = ';')
annotations$X9 <- gsub(".*:", "", annotations$X9)
annotations$X9 <- gsub(".*=", "", annotations$X9)
annotations$Name <- gsub(".*:", "", annotations$Name)
annotations$Name <- gsub(".*=", "", annotations$Name)

genes_list <- annotations %>% filter(X3 == "gene")

genes_list$length <- genes_list$X5 - genes_list$X4

## Contains the gene name and gene length
gene_length_df <- genes_list %>% select('Name', 'length')
```

## Format GeneInfo File

```{r format geneinfo}
geneinfo$gene <- stringr::str_replace_all(geneinfo$gene, "Gene:", "")
## Add in mitochondrial genes to gene info
geneinfo <- add_mitochondrial_genes(df = geneinfo, column_name = "gene")

## Fix NA Values
geneinfo$interpro_description[which(is.na(geneinfo$interpro_description))] <- "No InterPro entry"
geneinfo$go_biologicalprocess[which(is.na(geneinfo$go_biologicalprocess))] <- "No GO terms for biological process"
geneinfo$go_cellularcomponent[which(is.na(geneinfo$go_cellularcomponent))] <- "No GO terms for cellular component"
geneinfo$go_molecularfunction[which(is.na(geneinfo$go_molecularfunction))] <- "No GO terms for molecular function"
```

## Format Tidy Counts File into Counts Matrix

```{r format counts file}
## rRNAs that have been misannotated as mRNAs
genes_overlapping_with_rRNAs <- c("WBGene00228061", "WBGene00268654", "WBGene00268655", "WBGene00268656", "WBGene00268657", "WBGene00221211",
 "WBGene00220288", "WBGene00220284", "WBGene00268299", "WBGene00228060", "WBGene00268300")
colnames(counts.orig) <- c("Name", "counts", "replicate", "Project")
#counts.orig <- counts.orig %>% tidyr::separate("counts", into = c("counts", "stage"), sep = ' ')
counts.orig$counts <- as.double(counts.orig$counts)
```

```{r}
## Remove Gene: prefix from gene names
counts.orig$Name <- stringr::str_replace_all(counts.orig$Name, "Gene:", "")
counts.orig <- counts.orig %>% dplyr::arrange(replicate, Name)
## Remove __* lines generated from HTSeq-Counts
counts.orig <- counts.orig[!grepl("__", counts.orig$Name),]
## Remove misannotated mRNAs
counts.orig <- counts.orig[!(counts.orig$Name %in% genes_overlapping_with_rRNAs),]
## Remove all counts mapping to Wolbachia
counts.orig <- counts.orig[!(grepl("Wbm", counts.orig$Name)),] ## Remove Wolbachia counts

counts.orig <- dplyr::inner_join(counts.orig, gene_length_df, by = "Name") %>% arrange(replicate, Name)
```

## Convert counts data frame to counts matrix (row = gene, columns = replicate, values are counts)

```{r}
counts <- create_matrix(df = counts.orig, row = "Name", centre = "counts", column = "replicate")
```

```{r}
mRNA_features <- annotations %>% filter(X3 == "mRNA")
## Format names to isolate gene name
mRNA_features$X9 <- str_remove(mRNA_features$X9, "Parent=Gene:")
mRNA_features$Name <- gsub(".*:", "", mRNA_features$Name)
mRNA_features <- unique(mRNA_features$X9)

## Remove the mitochondrial genes from couints
counts.mito <- counts[grepl("gene-", rownames(counts)),]
## Select only genes in mRNAs list
counts.WB.mRNA <- counts[(rownames(counts) %in% mRNA_features),]
## Combine mRNA gene counts with mitochondrial gene counts
counts <- rbind(counts.mito, counts.WB.mRNA)
```
## Convert Counts Matrix into TPM

```{r convert to tpm}
## Make sure that the tidy counts dataframe has the same genes as the counts matrix
counts.orig <- counts.orig[(counts.orig$Name %in% rownames(counts)),] 

if(all(rownames(counts) == unique(counts.orig$Name))){ ## Makes sure the genes are in the same order for the correct division

  tpm <- calculate_TPM(counts.mat = counts, counts.orig = counts.orig)

}else{
  message('The genes are not in the same order')
}

counts <- counts[,levels(groups$Replicate)]
tpm <- tpm[,levels(groups$Replicate)]


```



## Find DE Genes
```{r}
find_de_genes <- function(groups.to.use, counts, grouping_factor, return_genes = FALSE, returnDGE = FALSE){
  groups.to.use$grouping_factor <- grouping_factor
  groups.to.use <- factor_groups(groups.to.use)
  counts.to.use <- counts[,colnames(counts) %in% groups.to.use$Replicate]
  # Calculate Differentially Expressed Genes
  if(returnDGE){
      y.export <- identify_de_genes(counts.mat = counts.to.use, groups.df = groups.to.use, returnDGE = TRUE)
  }else{
      degenes.to.use <- identify_de_genes(counts.mat = counts.to.use, groups.df = groups.to.use)
  }
  if(return_genes){
    return(degenes.to.use)
  }else if(returnDGE){
    return(y.export)
  }
}
```

## Generate Rarefaction Curve
```{r}
rarefy.counts <- round(counts,0)
raremax <- round(min(rowSums(t(rarefy.counts))),0)
srare <- vegan::rarefy(t(rarefy.counts),raremax)
rarefy.raw.df <- vegan::rarecurve(t(rarefy.counts), step = round(raremax/10,0), sample = raremax, tidy = T)
rarefy.raw.df <- rarefy.raw.df %>% group_by(Site)

rarefy.df <- as.data.frame(matrix(nrow = 0,
                                  ncol = 5))
rarefy.points.df <- rarefy.df
for(i in unique(rarefy.raw.df$Site)){
  steps <- as.numeric(rarefy.raw.df[rarefy.raw.df$Site == i,]$Sample)
  detected_genes <- as.numeric(rarefy.raw.df[rarefy.raw.df$Site == i,]$Species)
  
  rarefy.df <- as.data.frame(rbind(rarefy.df,
                                   cbind(as.numeric(steps),as.numeric(detected_genes),
                                         as.character(groups[groups$Replicate == i,]$Replicate),
                                         as.character(groups[groups$Replicate == i,]$Sample),
                                         as.character(groups[groups$Replicate == i,]$column_colour))))
  
  rarefy.points.df <- as.data.frame(rbind(rarefy.points.df,
                                          cbind(as.numeric(max(steps)),as.numeric(max(detected_genes)),
                                                as.character(groups[groups$Replicate == i,]$Replicate),
                                                as.character(groups[groups$Replicate == i,]$Sample),
                                                as.character(groups[groups$Replicate == i,]$column_colour))))
  
  
}

col <- as.character(groups$column_colour)
names(col) <- as.character(groups$Sample)



pdf(paste0(FIGURE_DIR, "/rarefaction_curve.pdf"),
     width = 10,
     height = 10)
  print(

ggplot()+
    geom_line(mapping=aes(x = as.numeric(as.character(rarefy.df$V1)), 
                          y = as.numeric(as.character(rarefy.df$V2)),
                          group = rarefy.df$V3,color = rarefy.df$V4)) +
    geom_point(mapping=aes(x = as.numeric(as.character(rarefy.points.df$V1)), 
                           y = as.numeric(as.character(rarefy.points.df$V2)),
                           group = rarefy.points.df$V3, color = rarefy.points.df$V4), size = 3) +
    scale_color_manual(values = col) +
    labs(x="number of reads mapping to features annotated as mRNA", y="number genes detected", color = "Sample") +
    theme_bw() + scale_y_continuous(expand = expansion(mult = c(0.2, .2))) + geom_hline(yintercept = nrow(counts)) 
  )

dev.off()

```


## Remove Single Replicate Samples
```{r}
single_replicate_samples <- as.data.frame(table(groups$Sample)) %>% filter(Freq == 1)


single_replicate <- groups[groups$Sample %in% single_replicate_samples$Var1,]
single_replicate <- factor_groups(single_replicate)

groups <- groups[!(groups$Sample %in% single_replicate_samples$Var1),]
groups <- factor_groups(groups)

counts <- counts[,!(colnames(counts) %in% single_replicate$Replicate)]
tpm <- tpm[,!(colnames(tpm) %in% single_replicate$Replicate)]

```

## Variance Analysis
```{r}

form <- ~ (1|Sample)
varPart.df.all <- variance_function(counts, groups, form)

pdf(paste0(FIGURE_DIR, "/all_samples_variancePartition_histogram.pdf"),
     width = 10,
     height = 10)
print(
 ggplot(varPart.df.all, aes(x = Variance_Explained)) + geom_histogram() + facet_wrap(~attribute, ncol = 1) + 
   labs(x = 'Variance explained ', title = 'Libro et al 2016')
)
dev.off()

pdf(paste0(FIGURE_DIR, "/all_samples_variancePartition.pdf"),
     width = 10,
     height = 10)
print(
  ggplot(varPart.df.all, aes(x = attribute, y = Variance_Explained * 100)) + 
  geom_boxplot() + labs(y = 'Variance explained (%)', title = 'Libro et al 2016')
)
dev.off()

```

## All Samples
```{r}

## 1 DE gene
degenes.all <- find_de_genes(groups, counts, groups$Sample, return_genes = TRUE)

y.all <- find_de_genes(groups, counts, groups$Sample, return_genes = FALSE, returnDGE = TRUE)

plot_de_genes(tpm = tpm, degenes = degenes.all, groups.df = groups, OUTFILE = "All_Samples_DE", column_colour = "column_colour")
plot_de_genes(tpm = tpm, degenes = degenes.all, groups.df = groups, OUTFILE = "All_Samples_log", column_colour = "column_colour", y_dge = y.all, wgcna_clustering = FALSE, z_score = FALSE)
```


```{r}

tpm.de <- tpm[rownames(tpm) %in% degenes.all,]
log2tpm.de <- log2(tpm.de + 1)
zscore.log2.tpm.de <- as.data.frame(t(scale(t(log2tpm.de))))

# WBGene00235408 ## snoRNA


```


```{r}
plot_variance_analysis <- function(counts.mat, groups.df, file_name, title_name){
  form <- ~ (1|Sample)
  groups.df <- factor_groups(groups.df)
  varPart.df <- variance_function(counts.mat[,colnames(counts.mat) %in% groups.df$Replicate], groups.df, form)
  
  pdf(paste0(FIGURE_DIR, "/", file_name, "_histogram.pdf"),
       width = 10,
       height = 10)
  print(
   ggplot(varPart.df, aes(x = Variance_Explained)) + geom_histogram() + facet_wrap(~attribute, ncol = 1) + 
     labs(x = 'Variance explained ', title = paste0(title_name))
  )
  dev.off()
  
  pdf(paste0(FIGURE_DIR, "/", file_name, ".pdf"),
       width = 10,
       height = 10)
  print(
    ggplot(varPart.df, aes(x = attribute, y = Variance_Explained * 100)) + 
    geom_boxplot() + labs(y = 'Variance explained (%)', title = paste0(title_name))
  )
  dev.off()

}
```



## Pairwise Differential Expression
```{r}

table_of_gene_number_in_heatmap <- tibble('Type_of_Comparison' = as.character() ,
                                          'Number_of_Genes_in_Heatmap' = as.numeric(),
                                          'Differential_Expression' = as.character())

pairwise_de_genes <- function(groups, counts, condition_1, condition_2, comparison, title, gene_number_table = table_of_gene_number_in_heatmap){
  groups <- groups %>% filter(grepl(condition_1, groups$Sample) | grepl(condition_2, groups$Replicate))
  degenes <- find_de_genes(groups, counts, groups$Sample, return_genes = TRUE)
  y <- find_de_genes(groups, counts, groups$Sample, return_genes = FALSE, returnDGE = TRUE)

  groups <- factor_groups(groups)
  if(length(degenes) < 3){
      tpm.keep <- tpm[rownames(tpm) %in% rownames(y$counts),colnames(tpm) %in% groups$Replicate]
      plot_pca(tpm = tpm.keep, groups.df = groups, OUTFILE = paste0("PCA_", comparison), counts = counts)
      groups <- plot_dendrogram(tpm = tpm.keep, groups.df = groups, OUTFILE = paste0("dendrogram_", comparison))

      plot_de_genes(tpm = tpm, degenes = degenes, groups.df = groups, OUTFILE = paste(comparison, "_log"), column_colour = "column_colour",
                y_dge = y, wgcna_clustering = FALSE, z_score = FALSE)
      plot_de_genes(tpm = tpm, degenes = degenes, groups.df = groups, OUTFILE = paste(comparison), column_colour = "column_colour",
                y_dge = y, wgcna_clustering = FALSE, z_score = TRUE)

      gene_number_table <- gene_number_table %>%
        add_row(Type_of_Comparison = paste0(comparison),
          Number_of_Genes_in_Heatmap = nrow(tpm.keep),
          Differential_Expression = "N")

  }else{

      tpm.de <- tpm[rownames(tpm) %in% degenes,colnames(tpm) %in% groups$Replicate]
      plot_pca(tpm = tpm.de, groups.df = groups, OUTFILE = paste0("PCA_", comparison), counts = counts)
      groups <- plot_dendrogram(tpm = tpm.de, groups.df = groups, OUTFILE = paste0("dendrogram_", comparison))


      plot_de_genes(tpm = tpm, degenes = degenes, groups.df = groups, OUTFILE = paste(comparison), column_colour = "column_colour")
      plot_de_genes(tpm = tpm, degenes = degenes, groups.df = groups, OUTFILE = paste(comparison, "_log"), column_colour = "column_colour",
                y_dge = y, wgcna_clustering = FALSE, z_score = FALSE)

      gene_number_table <- gene_number_table %>%
        add_row(Type_of_Comparison = paste0(comparison),
          Number_of_Genes_in_Heatmap = nrow(tpm.de),
          Differential_Expression = "Y")

  }

  counts <- counts[,levels(groups$Replicate)]
  plot_variance_analysis(counts = counts, groups = groups, file_name = paste0(comparison, "_variancePartition"), title_name = paste0(title))


  return(gene_number_table)
}

## No CnoA.36h provided
## No 16h dsDNA

## CvsA.24h 0 DE
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.24h", "CnoA.24h", "CplusA_vs_CnoA", "CplusA vs CnoA Samples")

## 22h EC vs CplusA.24h ## 173 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.24h", "Ec.22h", "CplusA24h_vs_Ec22h", "CplusA24h vs Ec22h Samples")
## 24h Ba vs CplusA.24h ## 463 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.24h", "Ba.24h", "CplusA24h_vs_Ba24h", "CplusA24h vs Ba 24h Samples")
## 36h Ba vs CplusA.36h ## 0 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.36h", "Ba.36h", "CplusA36h_vs_Ba36h", "Cplus36h vs Ba 36h Samples")
## 36h dsRNA vs CplusA.36h ## 0 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.36h", "RNA.36h", "CplusA36h_vs_RNA36h", "Cplus36h vs RNA 36h Samples")
## 12h dsDNA vs CplusA.24h ## 18 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.24h", "DNA.12h", "CplusA24h_vs_DNA12h", "CplusA24h vs DNA12h Samples")
## 24h dsRNA vs CplusA.24h ## 1 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.24h", "RNA.24h", "CplusA24h_vs_RNA24h", "CplusA24h vs RNA24h Samples")
## 20h EC vs CplusA.24h ## 30 DE genes
table_of_gene_number_in_heatmap <- pairwise_de_genes(groups, counts, "CplusA.24h", "Ec.20h", "CplusA24h_vs_Ec20h", "CplusA24h vs Ec20h Samples")


```


## Plotting Immune Genes
```{r}
immune_genes <- c("WBGene00223742", "WBGene00224039", "WBGene00225829", "WBGene00223030", "WBGene00223031", "WBGene00225025", "WBGene00221438",
                  "WBGene00222060", "WBGene00223763", "WBGene00226266", "WBGene00227305", "WBGene00227386", "WBGene00227253", "WBGene00226493",
                  "WBGene00227849", "WBGene00225033", "WBGene00223471", "WBGene00226742", "WBGene00220358", "WBGene00224494", "WBGene00226194",
                  "WBGene00225119", "WBGene00222209", "WBGene00223279", "WBGene00227737", "WBGene00227118", "WBGene00227744", "WBGene00225423",
                  "WBGene00226480", "WBGene00221940", "WBGene00222395", "WBGene00227909", "WBGene00222860", "WBGene00222412", "WBGene00225675",
                  "WBGene00225404", "WBGene00226691", "WBGene00223487", "WBGene00227131", "WBGene00226268", "WBGene00227825", "WBGene00222335",
                  "WBGene00227681", "WBGene00224538", "WBGene00226988", "WBGene00227595", "WBGene00223824", "WBGene00225962", "WBGene00226345", 
                  "WBGene00224243", "WBGene00227503", "WBGene00225328", "WBGene00226871", "WBGene00226337", "WBGene00226265", "WBGene00227560",
                  "WBGene00226851", "WBGene00227250")

counts.immune.genes <- counts[rownames(counts) %in% immune_genes,]
tpm.immune.genes <- tpm[rownames(tpm) %in% immune_genes,]
## No DE genes
degenes.immune.genes <- find_de_genes(groups, counts.immune.genes, groups$Sample, return_genes = TRUE)
plot_de_genes(tpm = tpm, degenes = immune_genes, groups.df = groups, OUTFILE = "Immune_Genes_Zscore_log", column_colour = "column_colour")




```




